Cellular Automata and Computational Universality

### 2018-08-14 paper/paper.tex
Fixed double free in simulation, updated paper

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}

@@

    Consider the cellular automaton defined by the simple rule:
    \begin{equation}
        \label{rule90}
        a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2}
    \end{equation}

@@

    \begin{figure}[h]
        \begin{center}
            \includegraphics[width=4in]{rule1.png}
            \caption{Pattern generated from applying Eq. 1 over 129 generations}
            \label{fig:rule90}
        \end{center}
    \end{figure}

    \subsection{Self-similarity}
    Figure \ref{fig:rule90} shows a visualisation of Eq. \ref{rule90}. A cell is
    filled white if it has the value 1, and black otherwise. Note that each new
    ``line'' of the visualisation represents the automaton in the next
    successive time-step from the one before; this automaton is one-dimensional,
    and so we may represent it in its entirety as a single row of cells. This
    pattern exhibits two important characteristics, the first of which is
    ``self-similarity'', meaning that portions of the pattern generated are
    indistinguishable from the whole when magnified. As Wolfram states, in this
    way ``the pattern is therefore invariant under rescaling... and may be
    characterised by a fractal dimension''\cite{WolframFractal}. One such
    definition of this is the Hausdorff-Besicovitch dimension\cite{Hausdorff}.

@@

    \begin{equation}
        \label{sierpinskiDim}
        \begin{split}
            \tfrac{1}{3} & = (\tfrac{1}{2})^D \\
            \log{\tfrac{1}{3}} &= D \log{\tfrac{1}{2}} \\
            D &= \tfrac{\log{3}}{\log{2}} \approx 1.585
        \end{split}
    \end{equation}

@@

    \subsection{Self-organisation}
    Second of the characteristics exhibited by this cellular automaton is
    ``self-organisation'',

\section{}
\section{Examples}
\section{Applications}
\section{Computational Universality}
\section{Firing Squad Synchronisation Problem}
\section{Majority Problem}
\section{}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}
\usepackage{diagbox}

@@

    Consider the cellular automaton defined by the simple rule:
    \begin{equation}
        \label{eq:rule90}
        a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2}
    \end{equation}

@@

    \begin{figure}[h]
        \begin{center}
            \includegraphics[width=4in]{rule90.png}
            \caption{Pattern generated from applying Eq. 1 over 129 generations}
            \label{fig:rule90}
        \end{center}
    \end{figure}

    \subsection{Self-similarity}
    Figure \ref{fig:rule90} shows a visualisation of Eq. \ref{eq:rule90}. A cell is
    filled white if it has the value 1, and black otherwise. Note that each new
    ``line'' of the visualisation represents the automaton in the next
    successive time-step from the one before; this automaton is one-dimensional,
    and so we may represent it in its entirety as a single row of cells. This
    pattern exhibits two important characteristics, the first of which is
    ``self-similarity'', meaning that portions of the pattern generated are
    indistinguishable from the whole when magnified. As Wolfram states, in this
    way ``the pattern is therefore invariant under rescaling... and may be
    characterised by a fractal dimension'' \cite{WolframFractal}. One such
    definition of this is the Hausdorff-Besicovitch dimension \cite{Hausdorff}.

@@

    \begin{equation}
        \label{sierpinskiDim}
        \begin{split}
            \tfrac{1}{3} &= (\tfrac{1}{2})^D \\
            \log{\tfrac{1}{3}} &= D \log{\tfrac{1}{2}} \\
            D &= \tfrac{\log{3}}{\log{2}} \approx 1.585
        \end{split}
    \end{equation}

@@

    \subsection{Self-organisation}
    Second of the characteristics exhibited by this cellular automaton is
    ``self-organisation'', whereby from an initial disorderly or chaotic state
    there appear ordered structures seemingly spontaneously. This is illustrated
    in Fig. \ref{fig:self-organisation} below: one can clearly see the emergence
    of similar triangular structures in the cone snail's shell, appearing from a
    chaotic background. This starting chaos is modelled in the cellular
    automaton by randomly assigning each cell a value of 0 or 1.

    \begin{figure}[h]%
        \centering
        \subfloat[Shell of a cone
        snail]{{\includegraphics[width=2in]{cone_shell.jpg}}}%
        \qquad
        \subfloat[Rule 90 with random starting
        configuration]{{\includegraphics[width=2in]{rule90_rand.png}}}%
        \caption{Natural processes can exhibit structures found in automata}
        \label{fig:self-organisation}
    \end{figure}

\section{Classifying automata}
    In order to classify automata, we can take into account both the size of the
    neighbourhood each cell considers going into the next generation, as well as
    the number of possible states in which a cell may be. Call these $N$ and $Q$
    respectively; in the case of Eq. \ref{eq:rule90}, $N=3$ and $Q=2$, and so the
    total number of different ``transitions'' from one state to the next is $N^Q
    = 8$. We may write out all of these transitions as follows:

    \begin{table}[h]
        \centering
        \begin{tabular}{c|c|c|c|c|c|c|c}
            111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
            \hline
            0   & 1   & 0   & 1   & 1   & 0   & 1   & 0
        \end{tabular}
    \end{table}

    Reading the bottom row of this transition table, we get the binary number
    $01011010_2 = 90_{10}$, allowing us to name this rule set Rule 90. Using
    this convention, we may now construct our own automata based solely on the
    knowledge of its name (at least for elementary cellular automata). For
    example, take Rule 110, introduced by Wolfram 1983 \cite{WolframRule110}. We
    first convert $110_{10}$ to binary, and then use this to fill in the table
    as above:

    \begin{table}[h]
        \begin{minipage}{.5\linewidth}
            \centering
            \begin{tabular}{c|c|c|c|c|c|c|c}
                111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
                \hline
                ? & ? & ? & ? & ? & ? & ? & ?
            \end{tabular}
        \end{minipage}%
        $ \Rightarrow $
        \begin{minipage}{.6\linewidth}
            \begin{tabular}{c|c|c|c|c|c|c|c}
                111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
                \hline
                0 & 1 & 1 & 0 & 1 & 1 & 1 & 0
            \end{tabular}
        \end{minipage}
    \end{table}

    To convert this to a concrete set of rules to simulate in a computer
    programs (in a more elegant way than a collection of eight \texttt{if}
    statements), we can put the information we have gathered so far in a
    Karnaugh map to get a Boolean algebraic expression of the ruleset, as in the
    table below. We use $A$ , $B$, and $C$ to denote $a_{i-1}$, $a_i$, and
    $a_{i+1}$, at time $t$, respectively. From this we can reduce the ruleset to
    the Boolean expression in Eq. \ref{eq:rule110} by observation.

    \begin{table}[h]
        %\caption{The Karnaugh map resulting from decoding Rule 110}
        %\label{kmap}
        \centering
        \begin{tabular}{|c||c|c|c|c|}\hline
            \diagbox{C}{AB} & 00 & 01 & 11 & 10 \\
            \hline
            0 & 0  & 1  & 1  & 0    \\
            \hline
            1 & 1  & 1  & 0  & 1    \\
            \hline
        \end{tabular}
    \end{table}

    \begin{equation}
        \label{eq:rule110}
        \begin{split}
            B^{t+1} &= \overline{A} \cdot C + B \cdot \overline{C} + \overline{B} \cdot C \\
            &= \overline{A} \cdot C + B \oplus C
        \end{split}
    \end{equation}

    The resulting visualisation in Fig. \ref{fig:rule110} exhibits similar
    characteristic of self-organisation, however it is hard to say whether it
    shows self-similarity as well.

    \begin{figure}[h]
        \begin{center}
            \includegraphics[width=4in]{rule110.png}
            \caption{Rule 110 generated using Eq. \ref{eq:rule110}}
            \label{fig:rule110}
        \end{center}
    \end{figure}

    \subsection{Elementary cellular automata}
    \subsection{Life-like cellular automata}
    \subsection{Cyclic cellular automata}

\section{Examples}
\section{Applications}
\section{Computational Universality}
\section{Firing Squad Synchronisation Problem}
\section{Majority Problem}
